---
title: "API Notes"
author: "Sydney Royal"
format: html
editor: visual
---

## APIs

- a way to automate pulling information from the web

**How it works:**

- send a request: looks like an R list

{ "city": "Santa Barbara", "temperature": 18.4, "humidity": 72, "conditions": "partly cloudy", "wind": { "speed": 3.2, "direction": "WSW" } }

- don't need to learn JSON, because R can turn this into a recognizable dataframe

- instructions are provided to make an API key, but you don't need it for this class

```{r}
# We'll use 2 packages 

# sending HTTP requests
#install.packages("httr2")

# parson JSON into R objects
#install.packages("jsonlite")

library(httr2)
library(jsonlite)
library(tidyverse)
```

### Standard API Workflow

- find the API

- build a request – use *request* and *req_url_query*

- send it

- parse the response

# Example 1: Air Quality

**API documentation:**

- important because it tells you the name of the things you can download – when you build a request you have to ask for exactly what you want to call.

- this is where you get the names of the data that you're going to build into your request

```{r}

# let's get hourly PM 2.5 data for Santa Barbara
# could automate the script to grab data from a site daily, so for example if you wanted air quality data for a single day, you would automate it instead of telling the user to download new data every day.

# Build the request URL
# can be either <- or =, doesn't matter 
# give it the name of the website
req <- request("https://air-quality-api.open-meteo.com/v1/air-quality")  %>%
  # this is the building of the query itself- what you want to get 
  req_url_query(
    # list some of the parameters that I need to give the API want
    # ask for data for a specific lat and long
    latitude  = 34.42,
    longitude = -119.70,
    # this particular API needs to be hourly data
    hourly    = "pm10",
    hourly = "pm2_5",
    # get it for the past 3 days
    past_days = 3       # last 3 days of data
  )

# what the variables for each request are going to be different for each API, but the API website will tell you what they are. 

# Send the request and parse the response
# save the response that's = to the request (again can either be <- or =), then it will return the request
resp <- req %>% req_perform()
# it returns stuff in a complicated JSON format

# what is this
# gives a little more information about it, and explains the response
attributes(resp)

str(resp)

# Convert  to an R list
# the simplifyVector to format NA's usually what you want

# gives you something that is more readable to R
# calling it aq_raw, build the response, put the body from the JSON structure into the response
aq_raw <- resp %>% resp_body_json(simplifyVector = TRUE)

# What did we get back?
names(aq_raw)
# now looks more like what we expect

# Step 2 — What's inside a key?
str(aq_raw$hourly)          # compact overview of structure
# you can see what's inside hourly

length(aq_raw$hourly$pm2_5) # how many values?



```

## **Make a data frame**

```{r}

# Extract what you need to a data frame
aq_df <- data.frame(
  # turn time into a POSIX data format
  
  # POSIX is a standardized way of representing time data
  # unlist takes it out of the list and takes out hourly and time 
  # says this is how you should interpret this date, then you can do calculations with this information and find the time the data was collected.
  time  = as.POSIXct(unlist(aq_raw$hourly$time), format = "%Y-%m-%dT%H:%M"),
  
  # again unlists and puts it into your dataframe
  pm10 = unlist(aq_raw$hourly$pm10),
  pm2_5 = unlist(aq_raw$hourly$pm2_5)
)

head(aq_df)

```

**Find the Mean PM10**

```{r}

pm_10_mean<- mean(aq_df$pm10)

pm_2_5_mean<- mean(aq_df$pm2_5)


```

## **Make a Graph**

```{r}
ggplot(aq_df, aes(x = time, y = pm2_5)) +
  geom_line(colour = "steelblue") +
  geom_hline(yintercept = 12, linetype = "dashed", colour = "orange") +
  annotate("text", x = min(aq_df$time), y = 13.5,
           label = "EPA annual standard (12 µg/m³)",
           hjust = 0, size = 3, colour = "orange") +
  labs(
    title = "PM2.5 Air Quality — Santa Barbara",
    x     = "Time",
    y     = "PM2.5 (µg/m³)"
  ) +
  theme_minimal()
```

## Adding variables you want

```{r}

# build a query in the same way we did before 
# you CANNOT have spaces after the comma
# same as before but with forecast days added

# Build the request URL
req <- request("https://air-quality-api.open-meteo.com/v1/air-quality")  %>%
  req_url_query(
    latitude  = 34.42,
    longitude = -119.70,
    # for the data in hourly you have to put the quotes with no spaces
    hourly    = "pm2_5,ozone",  # we can ask for multiple variables at once, you can't have spaces!!
    past_days = 3,  # last 3 days of data
    forecast_days = 6 # add forecase days
  )

# Send the request and parse the response
resp <- req %>% req_perform()

# what is this
attributes(resp)

# Convert  to an R list
aq2_raw <- resp %>% resp_body_json(simplifyVector = TRUE)

# What did we get back?
names(aq2_raw)

names(aq2_raw$hourly)

# Step 2 — What's inside a key?
names(aq_raw$hourly)          # compact overview of structure

# how many values?
length(aq2_raw$hourly$pm2_5) 

# puts ozone back into the list
length(aq2_raw$hourly$ozone)

# compare with previous call - longer time
# because we added forecast days
length(aq_raw$hourly$pm2_5)

```

**Turn into a dataframe again**

```{r}

# Extract what you need to a data frame
aq_df2 <- data.frame(
  # turn time into a POSIX data format
  time  = as.POSIXct(unlist(aq2_raw$hourly$time), format = "%Y-%m-%dT%H:%M"),
  pm2_5 = unlist(aq2_raw$hourly$pm2_5),
  # add ozone to the data frame
  ozone = unlist(aq2_raw$hourly$ozone)
)

head(aq_df2)

```

**Graph it**

```{r}

ggplot(aq_df2, aes(x = ozone, y = pm2_5)) +
  geom_point(colour = "steelblue") +
  labs(
    title = "Ozone vs PM2.5 Air Quality — Santa Barbara",
    x     = "Ozone (µg/m³)",
    y     = "PM2.5 (µg/m³)"
  ) +
  theme_minimal()

```

## Writing a Reusable API

```{r}

# if you want to get PM 2.5 specifically, you can make a function that asks for the lat and long for a certain number of days 
get_pm25 <- function(lat, lon, past_days = 7) {
  
  # Error checking -- makes sure the inputs are right
  ## | means or
  ## & means and -- therefore both conditions have to be true
  if (lat < -90 | lat > 90) {
    
    # this is the message you get 
    # can do stop("Latitude must be between -90 and 90.") -- to actually stop the running and tell it to give the user a warning
    return("Latitude must be between -90 and 90.")
  }
  if (lon < -180 | lon > 180) {
    return("Longitude must be between -180 and 180.")
  }                                                    # ← added this
  
  # there are 2 more functions in R called "stop" and "warning" 
  ## a clearer way to do error checking
  # stop just stops it from running
  ## you can do it with return, but sometimes it can get confused between that being the name of a function you're working with. 
  
  req <- request("https://air-quality-api.open-meteo.com/v1/air-quality") |>
    req_url_query(
      latitude  = lat,
      longitude = lon,
      
      # pm2_5 is manually put in by the user
      hourly    = "pm2_5",
      past_days = past_days
    )
  resp <- req |> req_perform()
  raw  <- resp |> resp_body_json(simplifyVector = TRUE)
  
  # converts it into a dataframe
  data.frame(
    lat   = lat,
    lon   = lon,
    time  = as.POSIXct(unlist(raw$hourly$time), format = "%Y-%m-%dT%H:%M"),
    pm2_5 = unlist(raw$hourly$pm2_5)
  )
}
# the last thing in the function is returned automatically 

```

## Can use the function with multiple locations

```{r}

# Define several CA coastal cities
locations <- data.frame(
  city = c("Santa Barbara", "Ventura", "San Luis Obispo"),
  lat  = c(34.42,           34.27,     35.28),
  lon  = c(-119.70,         -119.29,   -120.66)
)

# Use for loop to call our function for each row

for (i in 1:nrow(locations)) {
  city <- locations$city[i]
  lat  <- locations$lat[i]
  lon  <- locations$lon[i]
  
  # for each case it's going to run the function for the data and then puts it into a dataframe
  aq_df <- get_pm25(lat, lon, past_days = 3)
  aq_df$city <- city
  
  if (i == 1) {
    aq_all <- aq_df
  } else {
    aq_all <- bind_rows(aq_all, aq_df)
  }
}


# Daily averages by city
aq_daily <- aq_all %>%
  mutate(date = as.Date(time)) %>%
  group_by(city, date) %>%
  summarize(mean_pm2_5 = mean(pm2_5, na.rm = TRUE), .groups = "drop")

ggplot(aq_daily, aes(x = date, y = mean_pm2_5, colour = city)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(
    title  = "Daily Mean PM2.5 — Central Coast Cities",
    x      = "Date",
    y      = "PM2.5 (µg/m³)",
    colour = "City"
  ) +
  theme_minimal()

```
