---
title: "Control Structures Notes"
author: "Sydney Royal"
format: html
editor: visual
---

## Setup

```{r}
library(here)
library(tidyverse)

```

## Flow Control

- allowing your function to do different things depending on conditions

  - ex. IF then ELSE

IF ELSE

- a conditional statement that's TRUE or FALSE

```{r}
# if we make a =20, then we can determine something about a variable
a = 20
classr  = ifelse(a > 90, "high","low")
classr
```

You can add together conditionals

```{r}
# a is less than 90, and it's something else, | means OR 

a = 20
classr  = ifelse((a > 90) & (a < 10), "good", "bad")
classr


# | is OR 

a = 20
classr  = ifelse((a > 90) | (a < 10), "good", "bad")
classr

```

```{r}

# also works with multiple values
a = runif(min=1, max=100, n=5)
a
# then can rerun it and see 
classr  = ifelse((a > 90) | (a < 10), "good", "bad")
classr

```

```{r}
# ifelse can work with vectors
classr  = ifelse(a > 90, "high","low")
classr
```

```{r}
source(here("course-materials/R/summarize_diversity2.R"))

flowers = sample(c("rose","peony","daisy"), replace=T, prob = c(0.1, 0.8, 0.1),size=30)

summarize_diversity(flowers, index="B")$simpson
```

```{r}

# = assigns a value
# == tests whether something is equal to a value

# for example: flowers = sample ... 
## says that flowers is a specific thing

# but if I did === B
## then it finds if it fits in with what B is, it's not definite

```

## Multiple Conditions

```{r}
# creates 3 options by embedding an if statement 
# if a is greater than 10 (false, then low), if that's true is it greater than 50? if that's false, it's meduim, if it's true it's high


classr = ifelse((a>10), ifelse(a>50, "high","med"),"low")
```

```{r}

# what do you do if your choice isn't binary?
# case_when is a little bit cleaner to work with multiple ocnditions

# what you do without case_when
# fish are being graded
# steelhead are A
# tuna are B
# everything else is C

# with case_when, you give a list of conditions and then have a response

# fish are set as equal to steelhead
fish = "steelhead"

# first condition is if it's equal to steelhead ~ what we want it to be if it's steelhead, then if fish == "tuna" ~ make it B, if it's not tuna (!) AND (&) not steelhead, ~ then it's C
case_when((fish =="steelhead")  ~ "A", (fish =="tuna") ~ "B", !(fish == "tuna") & !(fish=="steelhead") ~ "C")


```

## Make this cleaner

Usually we should make a function in a separate R file, but for kicks we'll do it here

```{r}

# apply to all of our ocean
# start by making a function - this will be cleaner!

# function called classify fish and we want to know what type of fish it is
classify_fish = function(fish) {
  
  # between brackets is whatever you want the function to do, so just repeat the same case_when statement we just did
 class=  case_when((fish =="steelhead")  ~ "A", (fish =="tuna") ~ "B", !(fish == "tuna") & !(fish=="steelhead") ~ "C")
 
 # have it give you the class of the fish
 return(class) }

classify_fish("steelhead")

```

```{r}
# can now give it a whole ocean

ocean= sample(c("tuna","salmon","steelhead"))

classify_fish(ocean)


```

Designing a function

- what are the inputs and what are the outputs

Inputs

- nutrient concentration for at least 5 days

- temperature for at least 5 days

- thresholds for temperature and nutrient with default values

Outputs

- mean concentration

- pollution risk as low, med, or high

```{r}
source(here("R/classify_lake.R"))

# generate some data to try the function
temperature = runif(min=4, max=35, n=30)
nutrient = runif(min=5, max=40, n=30)

# try it
res=classify_lake(temperature=temperature, nutrient=nutrient)

# lets pick some numbers we know should give us a low value
nutrient = runif(min=0, max=5, n=30)
classify_lake(temperature=temperature, nutrient=nutrient)

```

## Error Checking

- something you do for the user of your function

```{r}

data = c("a","b")
mean(data)

# R returns a message because it can't make a mean of string data

# should give direction to the user to have some checks in the function that return messages
# helps you debug your own code

# having some error checking to make sure the input data is reasonable for example:
# if temperature < -100 or >100
# set results to NA or you could exit the function with a warning


```

## Example of error checking

```{r}

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  # make sure inputs are positive and if not set as NA so result will be NA
   if (height < 0) height=NA
  
 # an alterative is to exit the function prematurely - a better way to check instead of above
  if (flow < 0) return("Flow cannot be less than 0")
  #    return("Height cannot be less than 0")
   # if this ends up being true, then it just stops with this return
      
  # calculate power
    result = rho * height * flow * g * Keff
  return(result)
}

power_gen(2,3)

```
