---
title: "Modularity Notes"
author: "Sydney"
format: html
editor: visual
---

## Vectors and Functions

- the point of a function is to use it with multiple inputs

- vector arithmetic

## Example

```{r}
a= c(1,2,3)
b= c(3,4,6)
a+b

# sum is done for each element of the vector
# this vector airthmetic only works if the vectors are the same length
```

## We've been doing this

```{r}
# lots of situations where you can do vector arithmetic without a for loop

```

## Hydropower example

```{r}

# the function implements that model with the multiplied vectors
library(here)
library(tidyverse)
# source the function from a file - make sure you are in the right working directory
source(here("R/power_gen_orig.R"))
power_gen_orig(flow=2,height=5)


```

```{r}
# how we'd run it

power_gen_orig(flow=2, height = 5)


# could also make a bunch of different flows
# Vectors of flow and height
flow = runif(min=1, max=10, n=20)
height = runif(min=5, max =30, n=20)

power = power_gen_orig(flow, height)
power

# lots of functions will work with single and multiple values of data 
```

- you can have the same number of each of your inputs or just one of them that can be applied to all the others.

- so it can work with 1 value of Keff, but can't work with 2, because it doesn't know what to do with them.

- can get away without using a for loop sometimes, but not always.

## Example where you have to use a for loop

- NPV = net present value – a way of comparing when you have costs, or damages to handle in the future

  - creates a discount rate

```{r}
# asking for NPV for damages across the discount rates

# define inputs: interest rates and damages)

# write the function, and then use loops to fill in the structure


```

```{r}
# write a function to compute npv
source(here("course-materials/R/compute_NPV.R"))
compute_NPV


compute_NPV(20, discount=0.01, time=20)



# generate some input
damages = c(25,33,91,24)
damages




```

```{r}
# now use a nested for loop to populate
for (i in 1:length(damages)) { ////# thed fapp
  for (j in 1:length(discount_rates)) {
    npvs[i,j] = compute_NPV(value=damages[i], discount=discount_rates[j], time=yr)
  }
}
npvs
```

```{r}
* big picture- you get a list of damages and make for loop to n

% main progam-- main module
# main program that controls the flow

# be precise about what your gol is 



```

Inputs:

- what is the goal

- key building blocks or functions

- inputs and outputs

- how do you put the boxes together

  - after you make your flow chart, you implement as you go.

Example

- how much profit you will make by building

- flow chart should guide your project design
