---
title: "Class Notes Functions"
author: "Sydney Royal"
format: html
editor: visual
---

## Assignment

Steps

1.  Define the contract

- in a quarto document describe inputs, outputs and what the function will do

2.  Implement the function in R

- power_gen_orig is a good template to start with - copy and modify

- make sure you include documentation at the top for now follow style from power_gen_orig but you can be more descriptive if you want - as long as line starts with #, it is a comment and will not be executed as code

- save this file as a name.R file

- *hint* use factors to make it easier to count the number of individuals of each species

3.  Test the function with some example data that you generate

- include you test code in the quarto document

# Class Notes

## Package loading

```{r}
library(here)
library(ggplot2)
library(tidyverse)
```

## BOD Dataset

```{r}
# look at a R data set on Biological Oxygen Demand over time

# notice the use of data - R function that brings pre-stored datasets into the work space
data(BOD)

# notice "help" and the information that it provides
help(BOD)

# options for looking at data
View(BOD)

head(BOD)
```

## Sort Function

```{r}
# example of a function - SORT
result = sort(BOD[,"demand"], decreasing=TRUE, method="quick")
head(result)
```

Inputs determine "what"

Parameters determine "how"

Contracts:

- if the users give you A, the function will return with B

- specify A and B as precisely as you can - should include units, parameters, time, etc.

## Function format:

\# descriptions at the top of the inputs, outputs, and what the function does.

*Function Name* = function(inputs, parameters) {

body of the function (manipulation of inputs)

return(values to return)

}

- body= everything between {} - anything inside the body isn't shown and can be reused

- return= tells R what the output was

## Practice

```{r}
library(here)
# source the function from a file - make sure you are in the right working directory
source(here("R/power_gen_orig.R"))

# alternatively
# function definition in-line
power_gen_orig = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
result = rho * height * flow * g * Keff
return(result)
}

# note that height and flow don't have default values, so you must provide them when you use the function, but the other parameters have default values that you can overwrite if you want to


```

### How the function is used:

```{r}
# This use of the function
power_gen_orig(20,1)

```

- can get the same result from specifying the parts of the function you're calling– if these aren't specified, then R assumes it's the default order that they appear in the function and you have to memorize what order they're in.

```{r}
# is the same as this
power_gen_orig(height=20, flow=1)
```

- You can change the order of the names of the arguments if you don't want to have to worry about the order, just name the argument.

```{r}
# or this; where we use the names of the arguments to specify which is which - this is especially useful if you have a lot of arguments and don't want to remember the order
power_gen_orig(flow=1, height=20)
```

- you can also change the default settings– so Keff has a default value of 0.8, but can be set to be something else, but that must be specified.

```{r}
# change the defaults
# what if efficiency is less

power_gen_orig(20,1, Keff=0.2)
```

## Scoping

- defines where a variable can be seen

```{r}
# specifying function name without () shows code
# try it for internal R functions
power_gen_orig
```

```{r}
# use the function and save the results
reservoir1 = power_gen_orig(height=4, flow=7)

reservoir2 = power_gen_orig(height=4, flow=7, Keff=0.2)

# note that internal variables are not defined so these R statements will FAIL
# try it
#height
```

```{r}
#Keff

```

```{r}
#results
```

- Only if outputs are defined in your workspace "reservoir1 and reservoir2"

```{r}
# only the outputs are in your workspace
reservoir1
```

```{r}
reservoir2
```

## Reading in Functions

```{r}
library(here)
source(here("R/power_gen_orig.R"))
power_gen_orig(20,1)
```

## Testing the Function I made

```{r}

# make a dataset
plants <- data.frame(
  species = c("Woodrush",
              "Holly (seedlings)",
              "Bramble",
              "Yorkshire Fog",
              "Sedge"),
  
  count = c(2, 8, 1, 1, 3)
)

# view dataframe
plants


```

```{r}
# read in the function

source(here("R/biodiversity_function_royal.R"))
biodiv_index(plants$count, sum(plants$count))


```

# Functions Part 2

### Next Week

- We'll start building a project with multiple functions in it

### Returning Values in a Function

- return(value) to get back one value, and use lists to get more than one

Lists

- informal data structures in R – ex. could include a table where the number of rows is different for each column, can have numeric, characters, and factors all together

```{r}
# make a list

# trying to keep track of things happening at a market
# use keyword list to tell R we're making a list
# an element called id that's 2, there's a character variable that's high, and there's a content column with apple and cherry, and there's a cost


sale = list(id=2, quality="high", contents=c("apple","cherry"), cost=c(4,5))
sale


# gives you the same dstructure as your dataframe, but the columns don't have to have the same number of things

# how do we access parts of the list?
```

```{r}
#ways to access elements
sale$id

# using $ can get you access to the parts in the list

# if we were in a dataframe we would use df[1,2]

# you can do this sort of with a list by doing nameoflist[#item in the list]
sale[2]

# double square brackets makes sure that you get the data inside the element, while one [] is used for things that are not lists
sale[[4]]

# this shouldn't usually work, but it does
sale[4]

# say we want to access the 2nd fruit that was sold from the sale
sale$contents[2]

# could also do with like this
sale[[3]][2]


# and then by using the [[]] you can do things with the values inside the list
sale[[1]][[1]]*sale[[4]][[2]]

```

### Add something to a list

```{r}
# you can also just add anything to a list

sale$location = "famer's market"

sale
```

```{r}
 # build a function that returns that maximum and minumim of a list of numbers

# a function that returns the max and min of a vector of numbers

# make a function called minmax, it's gonna have vector thing in it, x. You'll get the minimum value of x and then you'll get the maximum value of x. Then you ask it to return the min and max with each of the values named
minmax = function(x) {
  minx = min(x)
  maxx = max(x)
  
  # if i didn't do min= max= then i wouldn't see those named as min and max
  return(list(min=minx, max=maxx))
}

minmax(c(3,5,2,8,1))

# if i made a 

```

## More complex example

```{r}
# Source the function
source(here("R/analyze_water_quality.R"))

# --- Use defaults (most common case) ---
results <- analyze_water_quality(ph = 7.2, dissolved_oxygen = 8.5, turbidity = 2.1)
results


# if i want a value to be different than the default I can overwrite it

# --- Override one limit, e.g. stricter turbidity standard for drinking water ---
results_strict <- analyze_water_quality(
  ph = 7.2, dissolved_oxygen = 8.5, turbidity = 2.1,
  turbidity_max = 1.0
)
results_strict
```

## Redo the biodiversity index to invert it

- regular index -1

```{r}

plants <- data.frame(
  species = c("Woodrush",
              "Holly (seedlings)",
              "Bramble",
              "Yorkshire Fog",
              "Sedge"),
  
  count = c(2, 8, 1, 1, 3)
)

# view dataframe
plants

# now inverse the function -- try from this side instead of from the function itself

source(here("R/biodiversity_function_royal.R"))
1-biodiv_index(plants$count, sum(plants$count))

```

```{r}

# I can make the probabilities different for a created sample -- just keep this in your back pocket
possible.vampires =c("lena", "dracula", "buffy", "ambrose", size=100, prob =c(0.05, 0.05, 0.07, 0.02), replace=TRUE)


# repeat with a list
source(here("R/biodiversity_function_royal.R"))

# lets compare two catches - with different levels of diversity
possible.fish =c("salmon","steelhead","shark","tuna","cod")

# we can control sample probablities as long as it sums to 1 - here we have a mix of fish but cod is most common

# to do this we assign probabities to each fish 
# high prob more likely to be sampled
# probabilities must sum to 1 
catch1 = sample(possible.fish, size=10, prob = c(0.2, 0.2, 0.1, 0.1, 0.4), replace=T)

# lets create a test case that should have low diversity, making most fish steelhead
catch2 = sample(possible.fish, size=10, prob = c(0.1, 0.7, 0.1, 0.1, 0.1), replace=T)

summarize_diversity

```
